---
title: "PA1_template.Rmd"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading and preprocessing data
First we should load the data to R studio 
```{r}
dat = read.csv('activity.csv', header = T)
head(dat)
```


Next, we should remove NA values by subseting our data and keep only the rows where each of the column has no NA values.

```{r}
clean <- subset(dat, (!is.na(dat$steps)) & (!is.na(dat$date)) & (!is.na(dat$interval)))
head(clean)
```
##What is mean total number of steps taken per day?
For making our histogram, first we should summarize the steps by day
```{r}

library(dplyr)
spd <- clean %>% group_by(date) %>% summarise( tsteps = sum(steps))
head(spd)

```

Then we are ready to make our first histogram and add the appropriate lines for mean and median

```{r}
hist(spd$tsteps, main = "Steps per day")

mean_spd <- round(mean(spd$tsteps), 1)
median_spd <- round(median(spd$tsteps), 1)

abline(v=mean_spd, lwd = 1, col = 'red')
abline(v=median_spd, lwd = 1, col = 'blue')

legend('topright', lty = 1, lwd = 2, col = c("red", "blue"), legend = c('Mean', 'Median'))

paste0("Mean Value is ",mean_spd," and Median Value is ", median_spd)


```
## What is average daily activity 
Our next goal is to plot the average daily activity pattern. Our first task is to aggregate the suitable measure as we did before
```{r}
aspd <- clean %>% group_by(interval) %>% summarise( tsteps = mean(steps))
head(aspd)
```

Then we can proceed in plotting our graph
```{r}
with(aspd, plot(interval, tsteps, type = 'l', main = 'Avg Steps by date', ylab = 'Average Steps'))
```
Lastly we can find the interval with the maximum number of steps as shown below
```{r}
maxstepsint <- aspd[which.max(aspd$tsteps), 1]
maxsteps <- aspd[which.max(aspd$tsteps), 2]
paste0('Max number of steps is contained in the following interval ', maxstepsint, ' with average number of steps ', round(maxsteps))
```
##Inputing Missing Values
First we have to report the number of missing values
```{r}
sum(is.na(dat$steps))
```
Now we have to replace the missing values
```{r}
fixeddat <- dat
replace = function(x, y){
    if(is.na(x))
    {
        return (y)  #replace
    }
    else return (x)  #no replace
}

fixeddat$steps = mapply(replace,fixeddat$steps, mean(fixeddat$steps, na.rm = TRUE))
```
Now we can create two histograms for comparing the differences
```{r}
org <- dat %>% group_by(date) %>% summarise( tsteps = sum(steps))
fixed <- fixeddat %>% group_by(date) %>% summarise( tsteps = sum(steps))

# orginal #

hist(org$tsteps, main = "Steps per day (Original Data)", xlab = "Total Steps per day")

mean_org <- round(mean(org$tsteps, na.rm = TRUE), 1)
median_org <- round(median(org$tsteps, na.rm = TRUE), 1)

abline(v=mean_org, lwd = 1, col = 'red')
abline(v=median_org, lwd = 1, col = 'blue')

legend('topright', lty = 1, lwd = 2, col = c("red", "blue"), legend = c('Mean', 'Median'))

paste0("Mean Value is ",mean_org," and Median Value is ", median_org)

#fixed #

hist(fixed$tsteps, main = "Steps per day (Fixed using Average)", xlab = "Total Steps per day")

mean_fixed <- round(mean(fixed$tsteps), 1)
median_fixed <- round(median(fixed$tsteps), 1)

abline(v=mean_fixed, lwd = 1, col = 'red')
abline(v=median_fixed, lwd = 1, col = 'blue')

legend('topright', lty = 1, lwd = 2, col = c("red", "blue"), legend = c('Mean', 'Median'))

paste0("Mean Value is ",mean_fixed," and Median Value is ", median_fixed)
```
We can see that the differences are minor. In fact the mean value is almost the same in both cases. Median value is increased by a little in the case of the fixed Values

##Are there differences in activity patterns between weekdays and weekends?
First we should distinguish the weekends from the res of the days
```{r}
weekday = function(x){
    if(x %in% c('Saturday', 'Sunday')) return('Weekend')
    else return('Weekday')
}
 
clean$day = weekdays(as.Date(clean$date))
clean$dcat = as.factor(apply(as.matrix(clean$day), 1, weekday))

stepsbydaytype <- clean %>% group_by(interval, dcat) %>% summarise( tsteps = mean(steps))

```
Finally we can plot the 2 charts in order to compare the activity patterns for these 2 categories. 
(For creating them we will use lattice package and we will try to replicate the plot shown in the example)

```{r}
library(lattice)
xyplot(tsteps ~ interval | dcat, data = stepsbydaytype,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```

For creating the images we used the following code for each case
```{r}
png('plot1.png', width=480, height=480)
dev.off()
```

